For Finitary Induction-Induction, \\ Induction is Enough

### 2020-05-01 paper/paper.tex
stuf
using AFS encoding

In this section we say what it means that a model of ETT supports
IITs. We first define the notion of IIT signature. Signatures for
algebraic theories are usually given by inductive definitions. On the
one hand, we take this even further: our notion of signature is given
by a small type theory tailer-made to describe signatures, we call
this the \emph{theory of IIT signatures}. On the other hand we would
like to avoid using a complicated inductive definition (a type theory
is a quotient inductive-inductive type) to describe a simpler class of
inductive types. Hence we use a Church encoding of the theory of IIT
signatures to define signatures thus avoiding the need for
pre-existing inductive definitions. Another feature of our signatures
is that as they can include types inside the model of ETT (such as
$\N$ in the $\mathsf{isEven}$--$\mathsf{isOdd}$), hence signatures
arex internal to the particular model of ETT.

We define the theory of IIT signatures by saying what its algebras
(models) are. We call the \emph{theory of IIT signatures algebras}
simply \emph{signature algebras}. The theory of signatures is a small
type theory consisting of a (1) a substitution calculus (category with
families, CwF \cite{Dybjer96internaltype}) equipped with (2) a
universe, (3) a function space where the domain is in the universe and
(4) another function space with external domain. We explain the usage
of these type formers through examples after the definitions.

\begin{definition}[Signature algebra, $\SignAlg$]\label{def:algebra}
  In a model of ETT, a signature algebra is an iterated $\Sigma$ type
  consisting of the following components: four (families of) sets, 17
  operations and 18 equalities.
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\
  & \Con && : \Set \\
  & \Ty  && : \Con\ra\Set \\
  & \Sub  && : \Con\ra\Con\ra\Set \\
  & \Tm  && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Set \\
  & \id && : \Sub\,\Gamma\,\Gamma \\
  & \blank\circ\blank && : \Sub\,\Theta\,\Delta\ra\Sub\,\Gamma\,\Theta\ra\Sub\,\Gamma\,\Delta \\
  & \ass && : (\sigma \circ \delta) \circ \nu = \sigma \circ (\delta \circ \nu) \\
  & \idl && : \id\circ\sigma = \sigma \\
  & \idr && : \sigma\circ\id = \sigma \\
  & \blank[\blank] && : \Ty\,\Delta\ra\Sub\,\Gamma\,\Delta\ra\Ty\,\Gamma \\
  & \blank[\blank] && : \Tm\,\Delta\,A\ra(\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma]) \\
  & [\id] && : A[\id] = A \\
  & [\circ] && : A[\sigma \circ \delta] = A[\sigma][\delta] \\
  & [\id] && : t[\id] = t \\
  & [\circ] && : t[\sigma \circ \delta] = t[\sigma][\delta] \\
  & \boldsymbol{\cdot} && : \Con \\
  & \epsilon && : \Sub\,\Gamma\,\boldsymbol{\cdot} \\
  & {\boldsymbol{\cdot}\eta} && : (\sigma : \Sub\,\Gamma\,\boldsymbol{\cdot}) \ra \sigma = \epsilon \\
  & \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  & \blank,\blank && : (\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma])\ra\Sub\,\Gamma\,(\Delta\ext A) \\
  & \pi_1 && : \Sub\,\Gamma\,(\Delta\ext A)\ra \Sub\,\Gamma\,\Delta \\
  & \pi_2 && : (\sigma : \Sub\,\Gamma\,(\Delta\ext A))\ra \Tm\,\Gamma\,(A[\pi_1 \sigma]) \\
  & {\pi_1\beta} && : \pi_1(\sigma, t) = \sigma \\
  & {\pi_2\beta} && : \pi_2(\sigma, t) = t \\
  & {\pi\eta} && : (\pi_1\,\sigma, \pi_2\,\sigma) = \sigma \\
  & {,\circ} && : (\sigma, t)\circ\delta = (\sigma\circ\delta, t[\delta]) \\
  & \rlap{(2) Universe} \\
  & \U && : \Ty\,\Gamma \\
  & \El && : \Tm\,\Gamma\,\U\ra\Ty\,\Gamma \\
  & {\U[]} && : \U[\sigma] = \U \\
  & {\El[]} && : (\El\,a)[\sigma] = \El\,(a[\sigma]) \\
  & \rlap{(3) Inductive parameters} \\
  & \Pi && : (a:\Tm\,\Gamma\,\U)\ra\Ty\,(\Gamma\ext\El\,a)\ra\Ty\,\Gamma \\
  & \blank\oldapp\blank && : \Tm\,\Gamma\,(\Pi\,a\,B)\ra (u : \Tm\,\Gamma\,(\El\,a))
  \ra \Tm\,\Gamma\,(\El\,(B[\id,\,u])) \\
  & {\Pi[]} && : (\Pi\,a\,B)[\sigma] = \Pi\,(a[\sigma])\,(B[\sigma\circ\p,\q]) \\
  & {\oldapp[]} && : (t\oldapp\alpha)[\sigma] = (t[\sigma])\mathop{\oldapp}(\alpha[\sigma]) \\
  & \rlap{(4) External parameters} \\
  & \Pim && : (T:\Set)\ra(T\ra\Ty\,\Gamma)\ra\Ty\,\Gamma \\
  & \blank\appm \blank && : \Tm\,\Gamma\,(\Pim\,T\,B)\ra(\alpha:T) \ra\Tm\,\Gamma\,(B\,\alpha) \\
  & {\Pim[]} && : (\Pim\,T\,B)[\sigma] = \Pim\,T\,(\lambda \alpha.(B\,\alpha)[\sigma]) \\
  & {\appm[]} && : (t\appm\alpha)[\sigma] = (t[\sigma])\mathop{\appm}\alpha
\end{alignat*}
Given an $M:\SignAlg$, we denote its components as $\Con^M$, $\Ty^M$,
$\Sub^M$, $\Tm^M$, $\id^M$, and so on.
\end{definition}

@@

\begin{definition}[IIT signature]
  An IIT signature is a context in an arbitrary signature algebra,
  that is, a function $(M:\SignAlg)\ra\Con^M$.
\end{definition}
This can be seen as a Church encoding of $\Con$. For our purposes it
is not a problem that signatures don't live at the lowest universe
level.

@@

\begin{example}[Example signatures]\label{ex:signatures}
  The signature for natural numbers is given by the function
  \[
    nat := \lambda M.(\boldsymbol{\cdot}^M\,\ext^M\,\U^M\,\ext^M\,\El^M\,0^M\,\ext^M\,1^M \Ra^M \El^M\,1^M).
  \]
  For readability we write the same signature without the $\lambda$
  and the superscripts below on the left and we also write an informal
  version with variable names on the right for comparison.
  \[
  \boldsymbol{\cdot}\,\ext\,\U\,\ext\,\El\,0\,\ext\,1 \Ra \El\,1 \hspace{5.9em}
  \boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \Ra \El\,N
  \]
  This signature is given by a context of length three. We start with
  the empty context $\boldsymbol{\cdot}$, then we declare a sort by
  $\U$, then we declare an operator producing an element of $\El\,0$
  where $0$ is a de Bruijn index referring to the sort $\U$. Finally
  we declare an operator which takes as input an element of the sort
  $1$ and produces an element of the same sort. Note the assymetry of
  the function type $\Ra$: on the left there needs to be an element of
  $\U$, while on the right any type can appear (including another
  function type). This ensures strict positivity of the operators.

  Lists with elements of a given $T : \Set$ type are given by the
  following signature. Here we use the function space with external
  domain $\Ram$ to include a $T$ in the signature. Again, we omit the
  $\lambda$ and the superscripts for readability.
  \begin{alignat*}{10}
    & \boldsymbol{\cdot}\,\ext\,\U\,\ext\,\El\,0\,\ext\,T\Ram 1\Ra \El\,1 \hspace{3.56em} \boldsymbol{\cdot}\,\ext\,L:\U\,\ext\,nil : \El\,L\,\ext\,cons : T\Ram L\Ra \El\,L
  \end{alignat*}

@@

We first explain it by columns: the first column ($^A$ components)
corresponds to the standard model (set model, metacircular
interpretation \cite{ttintt}): contexts are sets, types are families,
terms are functions, the universe $\U$ is given by $\Set$, function
spaces are given by the external function space. The $^\D$ column is a
logical predicate interpretation, $^\A$ and $^\D$ together are a unary
version of the parametric model for dependent types
\cite{10.1145/2535838.2535852}. Contexts are predicates, types are
families of predicates, terms say that the $^\A$ interpretation
respects the predicates (this is ususally called fundamental lemma of
the logical predicate). $\U$ is given by predicate space, the
predicate at a $\Pi$ type holds for a function if it respects the
predicates. For $\Pim$, the predicate is defined pointwise. The last
column ($^\S$) is a modified dependent logical relation which refers
to both $^A$ and $^\D$. Contexts are binary relations where the second
parameter depends on the first one, types are dependent variants of
this, terms say that the relation is respected by $^\A$ and $^\D$,
respectively. $\U$ is however not relation space, but a function and
$(\El\,a)^\S$ is the graph of the function $a^\S$. $\Pi^\S$ for a
function again says that the function respects the relation, however
we don't simply say
\[
  (\Pi\,a\,B)^\S\,\gamma^S\,f\,f^D := (\alpha:a^\A\,\gamma)(\alpha^D:a^\D\,\gamma^D\,\alpha)(\alpha^\S:(\El\,a)^\S\,\gamma^S\,\alpha\,\alpha^D)\ra B^\S\,\dots,
\]
as $(\El\,a)^\S\,\gamma^S\,\alpha\,\alpha^D$ is just an equality
$a^\S\,\gamma^S\,\alpha = \alpha^D$ which we can singleton
contract. So we omit $\alpha^D$ and this equality as an input and
replace $\alpha^D$ by $a^\S\,\gamma^S\,\alpha$ in the definition.

When viewing $\ADS$ by rows, we can see that it is a part of the CwF
model of type theory \cite[Section
7.4]{Kaposi:2019:CQI:3302515.3290315}. In the CwF model, a context is
given by a CwF. From the category, we only have objects ($\Gamma^\A$),
and from the families, we have the families for types $\Gamma^\D$ and
terms $\Gamma^\S$. Types are the corresponding parts of displayed
CwFs, substitutions are parts of CwF morphisms, terms are parts of CwF
sections. $\U$ is part of the CwF of sets, $\El\,a$ is the part of the
discrete displayed CwF coming from $a$ (which is a CwF-morphism from
$\Gamma$ to the CwF of sets). $\Pi$ is given as by a dependent product
of displayed CwFs where it is essential that the domain is discrete,
$\Pim$ is the pointwise direct product.

\begin{definition}[A model of ETT supports IITs]\label{sec:hasiits}
  A model of ETT supports IITs if for any signature
  $f:(M:\SignAlg)\ra\Con^M$ there is a $\con_f: (f\,\ADS)^\A$ and an
  $\elim_f : (\gamma^D:(f\,\ADS)^\D\,\con_f)\ra
  (f\,\ADS)^\S\,\con_f\,\gamma^D$.
\end{definition}
In other words, for any signature, we have an algebra called $\con$
(constructors) and for any displayed algebra over the constructors, we
have a section (called the eliminator).

@@

  For the signature $nat$ of natural numbers in \Cref{ex:signatures},

@@

  Note that as we work in a model of ETT, the computation rules that
  are expected from the elimination principle are given up to the
  internal equality type, however by equality reflection we also get
  them as definitional equalities.

@@

The definition of morphism is determined by the notion of algebra, it
can be computed following \cite{large_inf_qiit}, but we include it
here for completeness.
\begin{definition}[Signature morphism]\label{def:morphism}
  A morphism between signature algebras $M$ and $N$ consists of four
  functions and 17 equalities expressing that the functions preserve
  the operations of the two algebras. We use the same naming as in
  \Cref{def:algebra} and use superscripts to denote which algebra is
  meant.
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\
  & \Con && : \Con^M && \ra\Con^N \\
  & \Ty  && : \Ty^M\,\Gamma && \ra\Ty^N\,(\Con\,\Gamma) \\
  & \Sub  && : \Sub^M\,\Gamma\,\Delta && \ra\Sub^N(\Con\,\Gamma)\,(\Con\,\Delta) \\
  & \Tm  && : \Tm^M\,\Gamma\,A && \ra\Tm^N\,(\Con\,\Gamma)\,(\Ty\,A) \\
  & \id && : \Sub\,\id^M && = \id^N \\
  & {\circ} && : \sigma\circ^M\delta && = \Sub\,\sigma\circ^N\Sub\,\delta \\
  & [] && : A[\sigma]^M && = \Ty\,A[\Sub\,\sigma]^N \\
  & [] && : t[\sigma]^M && = \Tm\,t[\Sub\,\sigma]^N \\
  & \boldsymbol{\cdot} && : \Con\,\boldsymbol{\cdot}^M && = \boldsymbol{\cdot}^N \\
  & \epsilon && : \Sub\,\epsilon^M && = \epsilon^N \\
  & \ext && : \Con\,(\Gamma\ext^M A) && = \Con\,\Gamma\ext^N\Ty\,A \\
  & , && : \Sub\,(\sigma,^M t) && = \Sub\,\sigma ,^N\Tm\,t \\
  & \pi_1 && : \Sub\,(\pi_1^M\,\sigma) && = \pi_1^N\,(\Sub\,\sigma) \\
  & \pi_2 && : \Tm\,(\pi_2^M\,\sigma) && = \pi_2^N\,(\Sub\,\sigma) \\
  & \rlap{(2) Universe} \\
  & \U && : \Ty\,\U^M && = \U^N \\
  & \El && : \Ty\,(\El^M\,a) && = \El^N\,(\Tm\,a) \\
  & \rlap{(3) Inductive parameters} \\
  & \Pi && : \Ty\,(\Pi^M\,a\,B) && = \Pi^N\,(\Tm\,a)\,(\Ty\,B) \\
  & \oldapp && : \Tm\,(t\mathbin{{\oldapp}^M} u) && = \Tm\,t\mathbin{{\oldapp}^N}\,\Tm\,u \\
  & \rlap{(4) External parameters} \\
  & \Pim && : \Ty\,(\Pim^M\,T\,B) && = \Pim^N\,T\,(\lambda\alpha.\Ty\,(B\,\alpha)) \\
  & \appm && : \Tm\,(t\mathbin{{\appm}^M}\alpha) && = \Tm\,t\mathbin{{\appm}^N}\alpha
\end{alignat*}
\end{definition}
\begin{definition}\label{def:theoryofsignatures}
  A model of ETT supports the theory of IIT signatures if there is a
  signature algebra equipped with a unique morphism into any
  algebra. We call this algebra the syntax or initial algebra, the
  morphism is called recursor.
\end{definition}

In this section, we first specify what it means that a model of ETT
supports all IITs. In order to even say this, the model already has to
support the theory of IIT signatures as given in
\Cref{def:theoryofsignatures}.

>>>>

In this section we say what it means that a model of ETT supports
IITs. We first define the notion of IIT signature. Signatures for
algebraic theories are usually given by inductive definitions. On the
one hand, we take this even further: our notion of signature is given
by a small type theory tailer-made to describe signatures, we call
this the \emph{theory of IIT signatures}. On the other hand we would
like to avoid using a complicated inductive definition (a type theory
is a quotient inductive-inductive type \cite{ttintt}) to describe a
simpler class of inductive types. Hence we use a Church encoding
\cite{DBLP:conf/lics/AwodeyFS18} of the theory of IIT signatures thus
avoiding the need for pre-existing inductive definitions. Another
feature of our signatures is that they can include types from the
model of ETT (such as $\N$ in the
$\mathsf{isEven}$--$\mathsf{isOdd}$). This is why signatures are
specified internally to the particular model of ETT.\footnote{There is
  another method inspired by Capriotti \cite{paolo} which allows
  stating what it means that any CwF $\mathcal{C}$ (not necessarily a
  model of ETT) supports IITs with definitional computation rules. In
  this method, signatures are described in the internal language of
  $\hat{\mathcal{C}}$, the presheaf model over $\mathcal{C}$. We don't
  use this approach because it is more technical, and it would not
  strengthen our main theorem (TODO: refer) as theorem (TODO: refer)
  needs $\mathcal{C}$ to be a model of ETT.}

We define the theory of IIT signatures by saying what its algebras
(models) are. We call the \emph{theory of IIT signatures algebras}
simply \emph{signature algebras}. The theory of signatures is a small
type theory consisting of a (1) a substitution calculus (category with
families, CwF \cite{Dybjer96internaltype}) equipped with (2) a
universe, (3) a function space where the domain is in the universe and
(4) another function space with external domain. We explain the usage
of these type formers through examples after the definition.

\begin{definition}[Signature algebra, $\SignAlg$]\label{def:algebra}
  In a model of ETT, a signature algebra is an iterated $\Sigma$ type
  consisting of the following four (families of) sets, 17 operations
  and 18 equalities.
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\
  & \Con && : \Set \\
  & \Ty  && : \Con\ra\Set \\
  & \Sub  && : \Con\ra\Con\ra\Set \\
  & \Tm  && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Set \\
  & \id && : \Sub\,\Gamma\,\Gamma \\
  & \blank\circ\blank && : \Sub\,\Theta\,\Delta\ra\Sub\,\Gamma\,\Theta\ra\Sub\,\Gamma\,\Delta \\
  & \ass && : (\sigma \circ \delta) \circ \nu = \sigma \circ (\delta \circ \nu) \\
  & \idl && : \id\circ\sigma = \sigma \\
  & \idr && : \sigma\circ\id = \sigma \\
  & \blank[\blank] && : \Ty\,\Delta\ra\Sub\,\Gamma\,\Delta\ra\Ty\,\Gamma \\
  & \blank[\blank] && : \Tm\,\Delta\,A\ra(\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma]) \\
  & [\id] && : A[\id] = A \\
  & [\circ] && : A[\sigma \circ \delta] = A[\sigma][\delta] \\
  & [\id] && : t[\id] = t \\
  & [\circ] && : t[\sigma \circ \delta] = t[\sigma][\delta] \\
  & \boldsymbol{\cdot} && : \Con \\
  & \epsilon && : \Sub\,\Gamma\,\boldsymbol{\cdot} \\
  & {\boldsymbol{\cdot}\eta} && : (\sigma : \Sub\,\Gamma\,\boldsymbol{\cdot}) \ra \sigma = \epsilon \\
  & \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  & \blank,\blank && : (\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma])\ra\Sub\,\Gamma\,(\Delta\ext A) \\
  & \pi_1 && : \Sub\,\Gamma\,(\Delta\ext A)\ra \Sub\,\Gamma\,\Delta \\
  & \pi_2 && : (\sigma : \Sub\,\Gamma\,(\Delta\ext A))\ra \Tm\,\Gamma\,(A[\pi_1 \sigma]) \\
  & {\pi_1\beta} && : \pi_1(\sigma, t) = \sigma \\
  & {\pi_2\beta} && : \pi_2(\sigma, t) = t \\
  & {\pi\eta} && : (\pi_1\,\sigma, \pi_2\,\sigma) = \sigma \\
  & {,\circ} && : (\sigma, t)\circ\delta = (\sigma\circ\delta, t[\delta]) \\
  & \rlap{(2) Universe} \\
  & \U && : \Ty\,\Gamma \\
  & \El && : \Tm\,\Gamma\,\U\ra\Ty\,\Gamma \\
  & {\U[]} && : \U[\sigma] = \U \\
  & {\El[]} && : (\El\,a)[\sigma] = \El\,(a[\sigma]) \\
  & \rlap{(3) Inductive parameters} \\
  & \Pi && : (a:\Tm\,\Gamma\,\U)\ra\Ty\,(\Gamma\ext\El\,a)\ra\Ty\,\Gamma \\
  & \blank\oldapp\blank && : \Tm\,\Gamma\,(\Pi\,a\,B)\ra (u : \Tm\,\Gamma\,(\El\,a))
  \ra \Tm\,\Gamma\,(\El\,(B[\id,\,u])) \\
  & {\Pi[]} && : (\Pi\,a\,B)[\sigma] = \Pi\,(a[\sigma])\,(B[\sigma\circ\p,\q]) \\
  & {\oldapp[]} && : (t\oldapp\alpha)[\sigma] = (t[\sigma])\mathop{\oldapp}(\alpha[\sigma]) \\
  & \rlap{(4) External parameters} \\
  & \Pim && : (T:\Set)\ra(T\ra\Ty\,\Gamma)\ra\Ty\,\Gamma \\
  & \blank\appm \blank && : \Tm\,\Gamma\,(\Pim\,T\,B)\ra(\alpha:T) \ra\Tm\,\Gamma\,(B\,\alpha) \\
  & {\Pim[]} && : (\Pim\,T\,B)[\sigma] = \Pim\,T\,(\lambda \alpha.(B\,\alpha)[\sigma]) \\
  & {\appm[]} && : (t\appm\alpha)[\sigma] = (t[\sigma])\mathop{\appm}\alpha
\end{alignat*}
Given an $M:\SignAlg$, we denote its components as $\Con^M$, $\Ty^M$,
$\Sub^M$, $\Tm^M$, $\id^M$, and so on.
\end{definition}

@@

The definition of morphism is determined by the notion of algebra, it
can be computed following \cite{large_inf_qiit}, but we include it
here for completeness.
\begin{definition}[Signature morphism, $\SignMor$]\label{def:morphism}
  A morphism from signature algebras $M$ to $N$ denoted
  $\SignMor\,M\,N$ consists of four functions and 17 equalities
  expressing that the functions preserve the operations of the two
  algebras. We use the same naming as in \Cref{def:algebra} and use
  superscripts to denote which algebra is meant.
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\
  & \Con && : \Con^M && \ra\Con^N \\
  & \Ty  && : \Ty^M\,\Gamma && \ra\Ty^N\,(\Con\,\Gamma) \\
  & \Sub  && : \Sub^M\,\Gamma\,\Delta && \ra\Sub^N(\Con\,\Gamma)\,(\Con\,\Delta) \\
  & \Tm  && : \Tm^M\,\Gamma\,A && \ra\Tm^N\,(\Con\,\Gamma)\,(\Ty\,A) \\
  & \id && : \Sub\,\id^M && = \id^N \\
  & {\circ} && : \sigma\circ^M\delta && = \Sub\,\sigma\circ^N\Sub\,\delta \\
  & [] && : A[\sigma]^M && = \Ty\,A[\Sub\,\sigma]^N \\
  & [] && : t[\sigma]^M && = \Tm\,t[\Sub\,\sigma]^N \\
  & \boldsymbol{\cdot} && : \Con\,\boldsymbol{\cdot}^M && = \boldsymbol{\cdot}^N \\
  & \epsilon && : \Sub\,\epsilon^M && = \epsilon^N \\
  & \ext && : \Con\,(\Gamma\ext^M A) && = \Con\,\Gamma\ext^N\Ty\,A \\
  & , && : \Sub\,(\sigma,^M t) && = \Sub\,\sigma ,^N\Tm\,t \\
  & \pi_1 && : \Sub\,(\pi_1^M\,\sigma) && = \pi_1^N\,(\Sub\,\sigma) \\
  & \pi_2 && : \Tm\,(\pi_2^M\,\sigma) && = \pi_2^N\,(\Sub\,\sigma) \\
  & \rlap{(2) Universe} \\
  & \U && : \Ty\,\U^M && = \U^N \\
  & \El && : \Ty\,(\El^M\,a) && = \El^N\,(\Tm\,a) \\
  & \rlap{(3) Inductive parameters} \\
  & \Pi && : \Ty\,(\Pi^M\,a\,B) && = \Pi^N\,(\Tm\,a)\,(\Ty\,B) \\
  & \oldapp && : \Tm\,(t\mathbin{{\oldapp}^M} u) && = \Tm\,t\mathbin{{\oldapp}^N}\,\Tm\,u \\
  & \rlap{(4) External parameters} \\
  & \Pim && : \Ty\,(\Pim^M\,T\,B) && = \Pim^N\,T\,(\lambda\alpha.\Ty\,(B\,\alpha)) \\
  & \appm && : \Tm\,(t\mathbin{{\appm}^M}\alpha) && = \Tm\,t\mathbin{{\appm}^N}\alpha
\end{alignat*}
Given an $f:\SignMor\,M\,N$, we denote its first four components just
by $f_\Con$, $f_\Ty$, $f_\Sub$, $f_\Tm$ or just write $f$ if it is
clear which one is meant.
\end{definition}

\begin{definition}[IIT signature]
  An IIT signature is a context in an arbitrary signature algebra
  which is compatible with morphisms:
  \begin{alignat*}{6}
    & \Sign :=\,\, && \big({sig}:(M:\SignAlg)\ra\Con^M\big)\times \\
    & && \big((M\,N:\SignAlg)(f:\SignMor\,M\,N)\ra f_\Con\,({sig}\,M) = {sig}\,N \big).
  \end{alignat*}
\end{definition}
The first component is the Church encoding of $\Con$, the second is
the compatibility condition introduced by Awodey, Frey and Speight
\cite{DBLP:conf/lics/AwodeyFS18}. We don't need impredicativity, as we
don't require that $\Sign$ is the initial signature algebra. The
compatibility condition says that if we obtain an $M$-context using
${sig}$ in model $M$ and then we transport it to $N$ using $f$, we get
the same $N$-context as directly applying ${sig}$ to $N$.

@@

\begin{example}[Example signatures]\label{ex:signatures}
  Using informal notation with variable names, the signature for
  natural numbers would be
  \[
    \boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \Ra \El\,N.
  \]
  This signature is a context of length three. We start with the empty
  context $\boldsymbol{\cdot}$, then we declare a sort $N:\U$, then we
  declare an operator producing an element of $N$ denoted by $\El\,N$,
  and finally we declare an operator which takes as input an element
  of the sort $N$ and produces an element of the same sort. Note the
  assymetry of the function type $\Ra$: on the left there needs to be
  an element of $\U$, while on the right any type can appear
  (including another function type). This ensures strict positivity of
  the operators.

  Our theory of signatures uses de Bruijn indices instead of variable
  names, so we rewrite the above signature to
  \[
    \boldsymbol{\cdot}\,\ext\,\U\,\ext\,z : \El\,0\,\ext\,s : 1 \Ra \El\,1,
  \]
  and as we $\Sign$ is defined formally as a pair of a Church encoding
  together with a compatibility condition, our final definition for
  the signature for natural numbers is the following. The
  compatibility condition is written using equational reasoning.
  \begin{alignat*}{5}
    & \rlap{$(nat,natc) := $} \\
    & \hspace{1em} && \big( && \rlap{$\lambda M.(\boldsymbol{\cdot}^M\,\ext^M\,\U^M\,\ext^M\,\El^M\,0^M\,\ext^M\,1^M \Ra^M \El^M\,1^M),$} \\
    & && && \lambda M\,N\,f\,.\, && f_\Con\,(\boldsymbol{\cdot}^M\,\ext^M\,\U^M\,\ext^M\,\El^M\,0^M\,\ext^M\,1^M \Ra^M \El^M\,1^M) = \\
    & && && && f_\Con\,(\boldsymbol{\cdot}^M\,\ext^M\,\U^M\,\ext^M\,\El^M\,0^M)\ext^N f_\Ty\,(1^N\Ra^N \El^N\,1^N) = \\
    & && && && f_\Con\,(\boldsymbol{\cdot}^M\,\ext^M\,\U^M)\,\ext^N\,f_\Ty\,(\El^M\,0^M)\ext^N f_\Tm\,1^N\Ra^M f_\Ty\,(\El^N\,1^N) = \\
    & && && && f_\Con\,\boldsymbol{\cdot}^M\,\ext^N\,f_\Ty\,\U^M\,\ext^N\,\El^N\,(f_\Tm\,0^M)\ext^N 1^M\Ra^M \El^M\,(f_\Tm\,1^N) = \\
    & && && && \boldsymbol{\cdot}^N\,\ext^N\,\U^N\,\ext^N\,\El^N\,0^N\,\ext^N\,1^N \Ra^N \El^N\,1^N \big)
  \end{alignat*}
  The first component builds the context describing natural numbers in
  $M$, the second one uses the fact that $f$ is a morphism, that is,
  it preserves all operations.

  Lists with elements of a given $T : \Set$ type are given by the
  following signature. Here we use the function space with external
  domain $\Ram$ to include a $T$ in the signature. For readability, we
  omit the $\lambda$ and the superscripts and we don't write the
  compatibility condition. On the right we list the same signature
  with variable names.
  \begin{alignat*}{10}
    & \boldsymbol{\cdot}\,\ext\,\U\,\ext\,\El\,0\,\ext\,T\Ram 1\Ra \El\,1 \hspace{3.56em} \boldsymbol{\cdot}\,\ext\,L:\U\,\ext\,nil : \El\,L\,\ext\,cons : T\Ram L\Ra \El\,L
  \end{alignat*}

@@

We first explain it by columns: the first column ($^\A$ components)
corresponds to the standard model (set model, metacircular
interpretation \cite{ttintt}): contexts are sets, types are families,
terms are functions, the universe $\U$ is given by $\Set$, function
spaces are given by the external function space. The $^\D$ column is a
logical predicate interpretation, $^\A$ and $^\D$ together are a unary
version of the parametric model for dependent types
\cite{10.1145/2535838.2535852}. Contexts are predicates, types are
families of predicates, terms say that the $^\A$ interpretation
respects the predicates (this is ususally called fundamental lemma of
the logical predicate). $\U$ is given by predicate space, the
predicate at a $\Pi$ type holds for a function if it respects the
predicates. For $\Pim$, the predicate is defined pointwise. The last
column $^\S$ is a modified dependent logical relation which refers to
both $^A$ and $^\D$. Contexts are binary relations where the second
parameter depends on the first one, types are dependent variants of
this, terms say that the relation is respected by $^\A$ and $^\D$,
respectively. $\U$ is however not relation space, but a function and
$(\El\,a)^\S$ is the graph of the function $a^\S$. $\Pi^\S$ for a
function again says that the function respects the relation, however
we don't simply say
\[
  (\Pi\,a\,B)^\S\,\gamma^S\,f\,f^D := (\alpha:a^\A\,\gamma)(\alpha^D:a^\D\,\gamma^D\,\alpha)(\alpha^\S:(\El\,a)^\S\,\gamma^S\,\alpha\,\alpha^D)\ra B^\S\,\dots,
\]
as $(\El\,a)^\S\,\gamma^S\,\alpha\,\alpha^D$ is just an equality
$a^\S\,\gamma^S\,\alpha = \alpha^D$ which we can singleton
contract. So we omit $\alpha^D$ and this equality as an input and
replace $\alpha^D$ by $a^\S\,\gamma^S\,\alpha$ in the definition.

When viewing $\ADS$ by rows, we can see that it is a part of the CwF
model of type theory \cite[Section
7.4]{Kaposi:2019:CQI:3302515.3290315}. In the CwF model, a context is
given by a CwF. Now, from the category part of the CwF, we only have
objects ($\Gamma^\A$), and from the families, we have the families for
types $\Gamma^\D$ and terms $\Gamma^\S$. Types are the corresponding
parts of displayed CwFs, substitutions are parts of CwF morphisms,
terms are parts of CwF sections. $\U$ is part of the CwF of sets,
$\El\,a$ is the part of the discrete displayed CwF coming from $a$
(which is a CwF-morphism from $\Gamma$ to the CwF of sets). $\Pi$ is
given by a dependent product of displayed CwFs where it is essential
that the domain is discrete, $\Pim$ is the pointwise direct product.

\begin{definition}[A model of ETT supports IITs]\label{sec:hasiits}
  A model of ETT supports IITs if for any signature $(sig,nat):\Sign$
  there is a
  \[
    \con_{sig}: ({sig}\,\ADS)^\A
  \]
  and an
  \[
    \elim_{sig} : (\gamma^D:({sig}\,\ADS)^\D\,\con_{sig})\ra ({sig}\,\ADS)^\S\,\con_{sig}\,\gamma^D.
  \]
\end{definition}
In other words, for any signature, we have an algebra called $\con$
(constructors) and for any displayed algebra over the constructors, we
have a section (called the eliminator).

@@

  For the signature $({nat},{natc})$ of natural numbers in \Cref{ex:signatures},

@@

  The computation rules of the elimination principle are only expected
  up to the internal equality type, but as we work with a model of
  ETT, by equality reflection we also get them as definitional
  equalities.

@@

In the previous section, using the notions of signature algebras and
signature morphisms, we defined signatures for IITs and what it means
for a model of ETT to support all IITs. In this section we show that
if a model of ETT supports the theory of IIT signatures, then it
supports all IITs.

The contents of this section are an adjustment of \cite[Sections 4 and
6]{Kaposi:2019:CQI:3302515.3290315} to our setting.

\begin{definition}\label{def:theoryofsignatures}
  A model of ETT supports the theory of IIT signatures if there is a
  signature algebra $\I : \SignAlg$ equipped with a unique morphism
  into any algebra $\rec : (M:\SignAlg)\ra\SignMor\,\I\,M$. We call
  this algebra the syntax or initial algebra, the morphism is called
  recursor.
\end{definition}
